---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# microdiluteR: analyze broth microdilution assays in R

![](https://img.shields.io/badge/R%20CMD%20check-pending-purple.svg)
[![Say Thanks](https://img.shields.io/badge/Thanks-message?label=Say&labelColor=%234b4b4b&color=%23ebfc03&link=https%3A%2F%2Fsaythanks.io%2Fto%2Fsilvia-eckert)](https://saythanks.io/to/silvia-eckert)


## :notebook: Background

The `microdiluteR` package is designed to help researchers tidy up data from photometer plates and provides functions to easily add metadata, regardless of whether the user is processing a single plate or multiple plates with complex metadata structures.

## :floppy_disk: Installation

You can install the development version of `microdiluteR` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("silvia-eckert/microdiluteR")
```

## :joystick: Usage

You can load `microdiluteR` as follows:

```{r example}
library(microdiluteR)
```


